Generating Random Logic Programs Using Constraint Programming

### 2020-05-30 text/paper2/paper.tex
New plots

For this experiment, we consider clauses of two types: \emph{rules} are clauses
such that the head atom has at least one variable, and \emph{facts} are clauses
with empty bodies and no variables. We use our constraint model to generate the
rules according to the following parameter values: $|\predicates{}|,
|\variables{}|, \maxNumNodes{} \in \{ 2, 4, 8 \}$, $\maxArity{} \in \{ 1, 2, 3
\}$, $\maxNumClauses{} = |\predicates{}|$, $\constants{} = \emptyset$. Just like
before, we explore all possible numbers of independent predicate pairs. We also
add a constraint that forbids empty bodies. For both rules and facts,
probabilities are uniformly sampled from $\{ 0.1, 0.2, \dots, 0.9 \}$.
Furthermore, all rules are probabilistic, while we vary the proportion of
probabilistic facts among \SI{25}{\percent}, \SI{50}{\percent}, and
\SI{75}{\percent}. For generating facts, we consider $|\constants{}| \in \{100,
200, 400 \}$ and vary the number of facts among $10^3$, $10^4$, and $10^5$ but
with one exception: the number of facts is not allowed to exceed
\SI{75}{\percent} of all possible facts with the given values of
$\predicates{}$, $\arities{}$, and $\constants{}$. Facts are generated using a
simple procedure that randomly selects a predicate, combines it with the right
number of constants, and checks whether the generated atom is already included
or not. We randomly select configurations from the description above and
generate ten programs with a complete restart of the constraint solver before
the generation of each program, including choosing different arities and
independent pairs. Finally, we set the query of each program to a random fact
not explicitly included in the program and consider six natively supported
algorithms and knowledge compilation techniques: binary decision diagrams (BDDs)
\cite{DBLP:journals/tc/Bryant86}, negation normal form (NNF), deterministic
decomposable NNF (d-DNNF) \cite{DBLP:journals/jair/DarwicheM02}, K-Best
\cite{DBLP:conf/pkdd/GutmannKKR08}, and two encodings based on sentential
decision diagrams \cite{DBLP:conf/ijcai/Darwiche11}, one of which encodes the
entire program (SDDX), while the other one encodes only the part of the program
relevant to the query (SDD).\footnote{Forward SDDs (FSDDs) and forward BDDs
  (FBDDs)
  \cite{DBLP:journals/corr/abs-1911-07750,DBLP:conf/ijcai/VlasselaerBKMR15} are
  omitted because the former uses too much memory and the implementation of the
  latter seems to be broken at the time of writing.}

@@

@@

%\begin{figure*}
%  \centering
%  \input{phase_transition.tex}
%  \caption{The distribution of the number of nodes in the binary search tree as
%    a function of the number of independent pairs of predicates for
%    $|\predicates{}| = 8$. Outliers are hidden, the dots denote mean values, and
%    the vertical axis is on a $\log_{10}$ scale.}
%  \label{fig:phase_transition}
%\end{figure*}

@@

% \section{Entailment Checking of Negative Cycles} \label{sec:cycles}

%\begin{algorithm}
%  \KwData{a program $\mathscr{P}$}
%  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
%  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
%  $\mathscr{P}$ with its $\variable{structure}$ and predicates in both body
%  and head fully determined\footnotemark\;
%  \lIf{\hasNegativeCycles($G_{\mathscr{R}}$)}{\Return{\textsc{false}}}
%  \leIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}{\Return{\textsc{undefined}}}
%  \caption{Entailment for negative cycles}
%  \label{alg:negative_cycles}
%\end{algorithm}
%\footnotetext{Variables and constants are irrelevant to our definition of
%  independence.}

% The algorithm takes all clauses whose structure and predicates have been fully
% determined and uses them to construct a full dependency graph. In our
% implementation, \texttt{hasNegativeCycles} function is just a simple extension
% of the backtracking cycle detection algorithm that `travels' around the graph
% following edges and checking if each vertex has already been visited or not.
% Alternatively, one could assign weights to the edges (e.g., $1$ for positive
% and $-\infty$ for negative edges), thus reducing our negative cycle detection
% problem to what is typically known as the negative cycle detection problem in
% the literature, and use an algorithm such as Bellman-Ford
% \cite{shimbel1954structure}.

% We checked that our model agrees with the derived combinatorial formula in up
% to a thousand different scenarios. The details of this empirical investigation
% are omitted as they are not crucial to the thrust of this paper.

>>>>

For this experiment, we consider clauses of two types: \emph{rules} are clauses
such that the head atom has at least one variable, and \emph{facts} are clauses
with empty bodies and no variables. We use our constraint model to generate the
rules according to the following parameter values: $|\predicates{}|,
|\variables{}|, \maxNumNodes{} \in \{ 2, 4, 8 \}$, $\maxArity{} \in \{ 1, 2, 3
\}$, $\maxNumClauses{} = |\predicates{}|$, $\constants{} = \emptyset$. Just like
before, we explore all possible numbers of independent predicate pairs. We also
add a constraint that forbids empty bodies. For both rules and facts,
probabilities are uniformly sampled from $\{ 0.1, 0.2, \dots, 0.9 \}$.
Furthermore, all rules are probabilistic, while we vary the proportion of
probabilistic facts among \SI{25}{\percent}, \SI{50}{\percent}, and
\SI{75}{\percent}. For generating facts, we consider $|\constants{}| \in \{100,
200, 400 \}$ and vary the number of facts among $10^3$, $10^4$, and $10^5$ but
with one exception: the number of facts is not allowed to exceed
\SI{75}{\percent} of all possible facts with the given values of
$\predicates{}$, $\arities{}$, and $\constants{}$. Facts are generated using a
simple procedure that randomly selects a predicate, combines it with the right
number of constants, and checks whether the generated atom is already included
or not. We randomly select configurations from the description above and
generate ten programs with a complete restart of the constraint solver before
the generation of each program, including choosing different arities and
independent pairs. Finally, we set the query of each program to a random fact
not explicitly included in the program and consider six natively supported
algorithms and knowledge compilation techniques: binary decision diagrams (BDDs)
\cite{DBLP:journals/tc/Bryant86}, negation normal form (NNF), deterministic
decomposable NNF (d-DNNF) \cite{DBLP:journals/jair/DarwicheM02}, K-Best
\cite{DBLP:conf/ijcai/RaedtKT07}, and two encodings based on sentential
decision diagrams \cite{DBLP:conf/ijcai/Darwiche11}, one of which encodes the
entire program (SDDX), while the other one encodes only the part of the program
relevant to the query (SDD).\footnote{Forward SDDs (FSDDs) and forward BDDs
  (FBDDs)
  \cite{DBLP:journals/corr/abs-1911-07750,DBLP:conf/ijcai/VlasselaerBKMR15} are
  omitted because the former uses too much memory and the implementation of the
  latter seems to be broken at the time of writing.}

@@

\appendix

% \section{Entailment Checking of Negative Cycles}

% \begin{algorithm}
%  \KwData{a program $\mathscr{P}$}
%  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
%  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
%  $\mathscr{P}$ with its $\variable{structure}$ and predicates in both body
%  and head fully determined\footnotemark\;
%  \lIf{\hasNegativeCycles($G_{\mathscr{R}}$)}{\Return{\textsc{false}}}
%  \leIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}{\Return{\textsc{undefined}}}
%  \caption{Entailment for negative cycles}
%  \label{alg:negative_cycles}
% \end{algorithm}
% \footnotetext{Variables and constants are irrelevant to our definition of
%  independence.}

% The algorithm takes all clauses whose structure and predicates have been fully
% determined and uses them to construct a full dependency graph. In our
% implementation, \texttt{hasNegativeCycles} function is just a simple extension
% of the backtracking cycle detection algorithm that `travels' around the graph
% following edges and checking if each vertex has already been visited or not.
% Alternatively, one could assign weights to the edges (e.g., $1$ for positive
% and $-\infty$ for negative edges), thus reducing our negative cycle detection
% problem to what is typically known as the negative cycle detection problem in
% the literature, and use an algorithm such as Bellman-Ford.

\section{Additional Plots}

\begin{figure*}
 \centering
 \input{phase_transition.tex}
 \caption{The distribution of the number of nodes in the search tree of the
   constraint solver as a function of the number of independent pairs of
   predicates for $|\predicates{}| = 8$. Outliers are hidden, the dots denote
   mean values, and the vertical axis is on a $\log_{10}$ scale. This plot
   suggests that exceptional runs when the constraint solver takes much longer
   than usual are most likely to occur with average levels of independence.}
 \label{fig:phase_transition}
\end{figure*}

\begin{figure}
  \centering
  \input{scatterplot.tex}
  \caption{An instance-by-instance comparison of the inference times of SDD and
    SDDX knowledge compilation techniques. The colours show the percentage of
    predicate pairs that were constrained to be independent. With
    mostly-independent programs, one would expect SDD to outperform SDDX,
    however, the data shows no significant difference.}
\end{figure}

%\begin{figure}
%  \centering
%  \includegraphics{algorithms.pdf}
%\end{figure}

%\begin{figure}
%  \centering
%  \includegraphics{cumulative.pdf}
%  \caption{The cumulative number of instances solved by inference algorithms for
%    increasing time limits}
%\end{figure}

\section{Example Programs}

In this appendix, we provide examples of probabilistic logic programs generated
by various combinations of parameters. In all cases, we use
\[
  [0.1, 0.2, \dots, 0.9, 1, 1, 1, 1, 1]
\]
as the list of probabilities. Each clause is written on a separate line and ends
with a full stop. The head and the body of each clause are separated with
\texttt{:-} (instead of $\gets$). The probability of each clause is prepended to
the clause, using \texttt{::} as a separator. Probabilities equal to one and
empty bodies of clauses can be omitted. Conjunction, disjunction, and negation
are denoted by commas, semicolons, and `\texttt{\textbackslash+}', respectively.
Parentheses are used to demonstrate precedence, although many of them are
redundant.

By setting $\predicates{} = [\texttt{p}]$, $\arities{} = [1]$, $\variables{} =
\{ \texttt{X} \}$, $\constants{} = \emptyset$, $\maxNumNodes{} = 4$, and
$\maxNumClauses{} = 1$, we get fifteen one-line programs, six of which are
without negative cycles (as highlighted below). Only the last program has no
cycles at all.

\begin{enumerate}[label=\arabic*.]
\item
\begin{verbatim}
0.5 :: p(X) :- (\+(p(X))), (p(X)).
\end{verbatim}
\item
\begin{verbatim}
0.8 :: p(X) :- (\+(p(X))); (p(X)).
\end{verbatim}
\hlitem
\begin{verbatim}
0.8 :: p(X) :- (p(X)); (p(X)).
\end{verbatim}
\hlitem
\begin{verbatim}
0.7 :: p(X) :- (p(X)), (p(X)).
\end{verbatim}
\item
\begin{verbatim}
0.6 :: p(X) :- (p(X)), (\+(p(X))).
\end{verbatim}
\item
\begin{verbatim}
p(X) :- (p(X)); (\+(p(X))).
\end{verbatim}
\hlitem
\begin{verbatim}
0.1 :: p(X) :- (p(X)); (p(X)); (p(X)).
\end{verbatim}
\hlitem
\begin{verbatim}
0.8 :: p(X) :- (p(X)), (p(X)), (p(X)).
\end{verbatim}
\item
\begin{verbatim}
p(X) :- \+(p(X)).
\end{verbatim}
\item
\begin{verbatim}
0.1 :: p(X) :- \+(\+(p(X))).
\end{verbatim}
\item
\begin{verbatim}
p(X) :- \+((p(X)); (p(X))).
\end{verbatim}
\item
\begin{verbatim}
0.4 :: p(X) :- \+((p(X)), (p(X))).
\end{verbatim}
\item
\begin{verbatim}
0.4 :: p(X) :- \+(\+(\+(p(X)))).
\end{verbatim}
\hlitem
\begin{verbatim}
0.7 :: p(X) :- p(X).
\end{verbatim}
\hlitem
\begin{verbatim}
p(X).
\end{verbatim}
\end{enumerate}

Note that:
\begin{itemize}
\item A program such as Program~14, because of its cyclic definition, defines a
  predicate that has probability zero across all constants. This can more easily
  be seen as solving the equation $0.7x = x$.
\item Programs 10 and 14 are not equivalent (i.e., double negation does not
  cancel out) because Program 10 has a negative cycle and is thus considered to
  be ill-defined.
\end{itemize}

To demonstrate variable symmetry reduction in action, we set $\predicates{} =
[\texttt{p}]$, $\arities{} = [3]$, $\variables{} = \{\texttt{X}, \texttt{Y},
\texttt{Z} \}$, $\constants{} = \emptyset$, $\maxNumNodes{} = 1$,
$\maxNumClauses{} = 1$, and forbid all cycles. This gives us the following five
programs:

\begin{itemize}
\item
\begin{verbatim}
0.8 :: p(Z, Z, Z).
\end{verbatim}
\item
\begin{verbatim}
p(Y, Y, Z).
\end{verbatim}
\item
\begin{verbatim}
p(Y, Z, Z).
\end{verbatim}
\item
\begin{verbatim}
p(Y, Z, Y).
\end{verbatim}
\item
\begin{verbatim}
0.1 :: p(X, Y, Z).
\end{verbatim}
\end{itemize}

This is one of many possible programs with $\predicates{} = [\texttt{p},
\texttt{q}, \texttt{r}]$, $\arities{} = [1, 2, 3]$, $\variables{} =
\{\texttt{X}, \texttt{Y}, \texttt{Z} \}$, $\constants{} = \{ \texttt{a},
\texttt{b}, \texttt{c} \}$, $\maxNumNodes{} = 5$, $\maxNumClauses{} = 5$, and
without negative cycles:

\begin{verbatim}
p(b) :- \+((q(a, b)), (q(X, Y)), (q(Z, X))).
0.4 :: q(X, X) :- \+(r(Y, Z, a)).
q(X, a) :- r(Y, Y, Z).
q(X, a) :- r(Y, b, Z).
r(Y, b, Z).
\end{verbatim}

Finally, we set
$\predicates{} = [\texttt{p}, \texttt{q}, \texttt{r}]$, $\arities{} = [1, 1,
1]$, $\variables{} = \emptyset$, $\constants{} = \{ \texttt{a} \}$,
$\maxNumNodes{} = 3$, $\maxNumClauses{} = 3$, forbid negative cycles, and
constrain predicates \texttt{p} and \texttt{q} to be independent. The resulting
search space contains thousands of programs such as:

\begin{itemize}
\item
\begin{verbatim}
0.5 :: p(a) :- (p(a)); (p(a)).
0.2 :: q(a) :- (q(a)), (q(a)).
0.4 :: r(a) :- \+(q(a)).
\end{verbatim}
\item
\begin{verbatim}
p(a) :- p(a).
0.5 :: q(a) :- (r(a)); (q(a)).
r(a) :- (r(a)); (r(a)).
\end{verbatim}
\item
\begin{verbatim}
p(a) :- (p(a)); (p(a)).
0.6 :: q(a) :- q(a).
0.7 :: r(a) :- \+(q(a)).
\end{verbatim}
\end{itemize}

@@

@@
